---
title: "HomeWork08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required packages

install.packages("quantmod")
install.packages("ellipse")
install.packages("PerformanceAnalytics")
install.packages("zoo")
install.packages("readxl")
install.packages("writexl")
```{r}
library(quantmod)
library(ellipse)
library(PerformanceAnalytics)   
library(zoo)
library(readxl)
library(writexl)
```

# Download and read the csv data

```{r}
FamaFrench = read.csv('FamaFrench_mon_69_98_3stocks.csv')
head(FamaFrench)
class(FamaFrench)
str(FamaFrench)
```

## Use multivariate regression and matrix algebra

```{r}
returns.mat = as.matrix(FamaFrench[, c(-1, -2,-3, -4, -5, -9 )])
market.mat = as.matrix(FamaFrench[,2, drop=F])
n.obs = nrow(returns.mat)
X.mat = cbind(rep(1,n.obs),market.mat)
colnames(X.mat)[1] = "intercept"
XX.mat = crossprod(X.mat)
```

# Multivariate Least Squares

```{r}
G.hat = solve(XX.mat)%*%crossprod(X.mat,returns.mat)
```

# Can also use solve(qr(X.mat), returns.mat)

```{r}
beta.hat = G.hat[2,]
E.hat = returns.mat - X.mat%*%G.hat
diagD.hat = diag(crossprod(E.hat)/(n.obs-2))
```

# Compute R2 values from multivariate regression

```{r}
sumSquares = apply(returns.mat, 2, function(x) {sum( (x - mean(x))^2 )})
R.square = 1 - (n.obs-2)*diagD.hat/sumSquares
R.square
```

# Print and plot results

```{r}
cbind(beta.hat, diagD.hat, R.square)
par(mfrow=c(1,2))
barplot(beta.hat, horiz=T, main="Beta values", col="blue", cex.names = 0.75, las=1)
barplot(R.square, horiz=T, main="R-square values", col="blue", cex.names = 0.75, las=1)
par(mfrow=c(1,1))
```

# Compute single index model covariance/correlation matrices

```{r}
cov.si = as.numeric(var(market.mat))*beta.hat%*%t(beta.hat) + diag(diagD.hat)
cor.si = cov2cor(cov.si)
```

# Compute global min variance portfolio
# Use single index covariance

```{r}
w.gmin.si = solve(cov.si)%*%rep(1,nrow(cov.si))
w.gmin.si = w.gmin.si/sum(w.gmin.si)
colnames(w.gmin.si) = "single.index"
```

# Use sample covariance

```{r}
w.gmin.sample = solve(var(returns.mat))%*%rep(1,nrow(cov.si))
w.gmin.sample = w.gmin.sample/sum(w.gmin.sample)
colnames(w.gmin.sample) = "sample"
cbind(w.gmin.si, sample = w.gmin.sample)
```

# Plot Single Index and Sample Weights

```{r}
par(mfrow=c(2,1))
barplot(t(w.gmin.si), horiz=F, main="Single Index Weights", col="blue", cex.names = 0.75, las=2)
barplot(t(w.gmin.sample), horiz=F, main="Sample Weights", col="blue", cex.names = 0.75, las=2)
par(mfrow=c(1,1))
```
